---
title: "Western Redcedar Point Defiance Data Analysis"
output: html_document
date: "2025-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Packages**

Load Tidyverse.
```{r}
library(tidyverse)
```

Load Stringdist.
```{r}
library(stringdist)
```

Load Dplyr.
```{r}
library(dplyr)
```

Load Tidytext.
```{r}
library(tidytext)
```

Load Fuzzyjoin.
```{r}
library(fuzzyjoin)
```

Load Permute (required package for vegan).
```{r}
library(permute)
```

Load Vegan.
```{r}
library(vegan)
```

# **Data Wrangling**

Import observations.
```{r}
i.nat.all <- read.csv("observations-601233.csv")
```

For the "Most Prevalent Understory Species" field our NMDS prep will consist of creating a list of species for filtering, setting everything to lowercase, CREATING FUCKING TOKENS WHAT THE FUCK IS THIS???

Vector to filter species prep for NMDS. (For this first NMDS run we will include "none" as a species.)
```{r}
SpeciesList <- c("huckleberry", "oregon grape", "himalayan blackberry", "trailing blackberry", "fern", "sword fern", "bracken fern", "salal", "salmon berry", "holly", "western starflower", "nettle", "fern", "moutain sweet cicely", "ocean spray", "pacific sanicle", "american trailplant", "horse chestnut", "cedar", "evergreen huckleberry", "wax myrtle", "stinking iris", "cotoneaster", "horsetail", "cherry laurel")
```

Lowercase everything, remove extra spaces, and common delimiters.
```{r}
Token.Prep <- i.nat.all %>%
  mutate(field.most.prevalent.understory.species = str_to_lower(field.most.prevalent.understory.species),
         field.most.prevalent.understory.species = str_replace_all(field.most.prevalent.understory.species, "(/| and |&|\\+),", " "),
         field.most.prevalent.understory.species = str_replace_all(field.most.prevalent.understory.species, ",", " "),
         field.most.prevalent.understory.species = str_squish(field.most.prevalent.understory.species))
```

Replace all unfilled observation fields with "none".
```{r}
Token.Prep <- Token.Prep %>%
  mutate(field.most.prevalent.understory.species = ifelse(field.most.prevalent.understory.species == "", "none", field.most.prevalent.understory.species))
```

Tokenize n = 1.
```{r}
Token.Prep.1gram <- Token.Prep %>%
  select(id, field.most.prevalent.understory.species) %>%
  unnest_tokens(output = "ngram", input = field.most.prevalent.understory.species, token = "ngrams", n = 1)
```

Tokenize n = 2.
```{r}
Token.Prep.2gram <- Token.Prep %>%
  select(id, field.most.prevalent.understory.species) %>%
  unnest_tokens(output = "ngram", input = field.most.prevalent.understory.species, token = "ngrams", n = 2)
```

Tokenize n = 3.
```{r}
Token.Prep.3gram <- Token.Prep %>%
  select(id, field.most.prevalent.understory.species) %>%
  unnest_tokens(output = "ngram", input = field.most.prevalent.understory.species, token = "ngrams", n = 3)
```

Merge all 3 token dataframes.
```{r}
Token.Prep.All <- bind_rows(Token.Prep.1gram, Token.Prep.2gram, Token.Prep.3gram) %>%
  distinct()
```

Fuzzy match.
```{r}
FuzzyMatchedSpeciesList <- stringdist_left_join(Token.Prep.All, 
  data.frame(ngram = SpeciesList), 
  method = "jw",
  max_dist = 0.15,
  distance_col = "dist")
```

Filter fuzzy match and remove NA.
```{r}
CleanedFuzzyMatch <- FuzzyMatchedSpeciesList %>%
  filter(!is.na(ngram.y))
```

Create species matrix.
```{r}
SpeciesMatrix <- CleanedFuzzyMatch %>%
  distinct(id, ngram.y) %>%
  mutate(presence = 1) %>%
  pivot_wider(names_from = ngram.y, values_from = presence, values_fill = 0)
```

Create new data frame of trees with tree canopy health data combined with Species Matrix.
```{r}
TreeCanopyHealthData <- i.nat.all[,c("id","field.tree.health.condition")]
  
CombinedMatrix <- left_join(TreeCanopyHealthData, SpeciesMatrix, by = "id")
```

We still have NA left over from observations that did not have any under story species observed, for our NMDS they need to be removed.
```{r}
CombinedMatrix2 <- na.omit(CombinedMatrix)
```

Then we remove all the observations with (No selection) for tree canopy health.
```{r}
CombinedCleanedMatrix <- CombinedMatrix2 %>% filter(field.tree.health.condition != "No selection")
```

Remove Tree Health Condition Column for NMDS and remove tree ids column.
```{r}
UnderstorySpeciesValuesMatrix <- CombinedCleanedMatrix %>%
  select(-id, -field.tree.health.condition)
```

Run NMDS.
```{r}
NMDSRun1 <- metaMDS(UnderstorySpeciesValuesMatrix,
                    distance = "jaccard",
                    binary = TRUE,
                    k = 2,
                    trymax = 10)
```

Checking results.
```{r}
print(NMDSRun1)
```

Checking orientation distance.
```{r}
stressplot(NMDSRun1)
```

Plotting graph.
```{r}
plot(NMDSRun1, display = "sites", type = "t")
```
Extract NMDSRun1 Scores.
```{r}
NMDSRun1Scores <- scores(NMDSRun1, display = "sites")
```

Separate the tree health scores column from our CombinedCleanMatrix Data Frame.
```{r}
TreeCanopyHealthScoringPrep <- CombinedCleanedMatrix[,c("field.tree.health.condition")]
```

Plot Graph.
```{r}
ggplot(NMDSRun1Scores, aes(x = NMDS1, y = NMDS2, color = TreeCanopyHealthScoringPrep)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Good" = "forestgreen", "Fair" = "yellow", "Poor" = "orange", "Dead" = "red")) +
  theme_minimal() +
  labs(title = "NMDS with Tree Health Scores",
       color = "Tree Health") +
  coord_cartesian(xlim = c(-3, 1), ylim = c(-.5, 1))
```

Bar chart setup.
```{r}
tcs=c("602", "128", "28")
```

```{r}
Tree.Condition.Summary <- data.frame(
  category=c("Good","Fair","Poor") , 
  value=as.numeric(tcs)
)
```

Load ggplot2
```{r}
library(ggplot2)
```

```{r}
ggplot(Tree.Condition.Summary, aes(x=reorder(category, -value) , y=value)) +
  geom_bar(stat = "identity") +
  labs(x = "Tree Condition", y = "Tree Count", title = "Tree Condition of Western Red Cedar Found at Point Defiance 2025")
```



For tree size a histogram with bucket size **5 inch** for now should suffice for our purposes.

Separate DBH column into its own data frame.
```{r}
DBH.Inch <- i.nat.all[c(12)]
```

Create histogram. (Bin = 5 Inches starting from 0)
```{r}
ggplot(DBH.Inch, aes(x = field.tree.diameter)) + 
  geom_histogram(binwidth = 5, boundary = 0, fill = "springgreen4", color = "black") + 
  stat_bin(binwidth = 5, boundary = 0, geom = "text", aes(label = after_stat(count), vjust = -0.5)) + 
  labs(x = "DBH", y = "Count", title = "Western Red Cedar Tree Sizes Found at Point Defiance 2025")
```

The 27 entries removed were entries without DBH's entered.

