---
title: "Western Redcedar Point Defiance Data Analysis"
output: html_document
date: "2025-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Packages/Libraries Used**

Load Tidyverse.
```{r}
library(tidyverse)
```

Load Stringdist.
```{r}
library(stringdist)
```

Load Dplyr.
```{r}
library(dplyr)
```

Load Tidytext.
```{r}
library(tidytext)
```

Load Fuzzyjoin.
```{r}
library(fuzzyjoin)
```

Load Permute (required package for vegan).
```{r}
library(permute)
```

Load Vegan.
```{r}
library(vegan)
```

# **Data Wrangling**

Import observations.
```{r}
i.nat.all <- read.csv("observations-601233.csv")
```

##NMDS Analysis Prep

For the "Most Prevalent Understory Species" field our NMDS prep will consist of creating a list of species for filtering, setting everything to lowercase and removing common delimiters, creating tokens, running a fuzzymatch based off of the tokens and seeing what matches up with our species filter, and creating a species matrix.

Vector to filter species prep for NMDS.
```{r}
SpeciesList <- c("huckleberry", "oregon grape", "himalayan blackberry", "trailing blackberry", "fern", "sword fern", "bracken fern", "salal", "salmon berry", "holly", "western starflower", "nettle", "moutain sweet cicely", "ocean spray", "pacific sanicle", "american trailplant", "horse chestnut", "cedar", "evergreen huckleberry", "wax myrtle", "stinking iris", "cotoneaster", "horsetail", "cherry laurel", "mountain huckleberry", "thimble berry", "ivy")
```

Lowercase everything, remove extra spaces, and common delimiters.
```{r}
Token.Prep <- i.nat.all %>%
  mutate(field.most.prevalent.understory.species = str_to_lower(field.most.prevalent.understory.species),
         field.most.prevalent.understory.species = str_replace_all(field.most.prevalent.understory.species, "(/| and |&|\\+),", " "),
         field.most.prevalent.understory.species = str_replace_all(field.most.prevalent.understory.species, ",", " "),
         field.most.prevalent.understory.species = str_squish(field.most.prevalent.understory.species))
```

Replace all unfilled observation fields with "none".
```{r}
Token.Prep <- Token.Prep %>%
  mutate(field.most.prevalent.understory.species = ifelse(field.most.prevalent.understory.species == "", "none", field.most.prevalent.understory.species))
```

Tokenize n = 1.
```{r}
Token.Prep.1gram <- Token.Prep %>%
  select(id, field.most.prevalent.understory.species) %>%
  unnest_tokens(output = "ngram", input = field.most.prevalent.understory.species, token = "ngrams", n = 1)
```

Tokenize n = 2.
```{r}
Token.Prep.2gram <- Token.Prep %>%
  select(id, field.most.prevalent.understory.species) %>%
  unnest_tokens(output = "ngram", input = field.most.prevalent.understory.species, token = "ngrams", n = 2)
```

Tokenize n = 3.
```{r}
Token.Prep.3gram <- Token.Prep %>%
  select(id, field.most.prevalent.understory.species) %>%
  unnest_tokens(output = "ngram", input = field.most.prevalent.understory.species, token = "ngrams", n = 3)
```

Merge all 3 token dataframes.
```{r}
Token.Prep.All <- bind_rows(Token.Prep.1gram, Token.Prep.2gram, Token.Prep.3gram) %>%
  distinct()
```

Fuzzy match.
```{r}
FuzzyMatchedSpeciesList <- stringdist_left_join(Token.Prep.All, 
  data.frame(ngram = SpeciesList), 
  method = "jw",
  max_dist = 0.15,
  distance_col = "dist")
```

Filter fuzzy match and remove NA.
```{r}
CleanedFuzzyMatch <- FuzzyMatchedSpeciesList %>%
  filter(!is.na(ngram.y))
```

Create species matrix.
```{r}
SpeciesMatrix <- CleanedFuzzyMatch %>%
  distinct(id, ngram.y) %>%
  mutate(presence = 1) %>%
  pivot_wider(names_from = ngram.y, values_from = presence, values_fill = 0)
```

Create new data frame of trees with tree canopy health data combined with Species Matrix.
```{r}
TreeCanopyHealthData <- i.nat.all[,c("id","field.tree.health.condition")]
  
CombinedMatrix <- left_join(TreeCanopyHealthData, SpeciesMatrix, by = "id")
```

We still have NA left over from observations that did not have any under story species observed, for our NMDS they need to be removed.
```{r}
CombinedMatrix2 <- na.omit(CombinedMatrix)
```

Then we remove all the observations with (No selection) for tree canopy health.
```{r}
CombinedCleanedMatrix <- CombinedMatrix2 %>% filter(field.tree.health.condition != "No selection")
```

Remove Tree Health Condition Column for NMDS and remove tree ids column.
```{r}
UnderstorySpeciesValuesMatrix <- CombinedCleanedMatrix %>%
  select(-id, -field.tree.health.condition)
```

##NMDS Analysis
Run NMDS.
```{r}
NMDSRun1 <- metaMDS(UnderstorySpeciesValuesMatrix,
                    distance = "jaccard",
                    binary = TRUE,
                    k = 2,
                    trymax = 10)
```

Checking results.
```{r}
print(NMDSRun1)
```

Checking orientation distance.
```{r}
stressplot(NMDSRun1)
```

Plotting graph.
```{r}
plot(NMDSRun1, display = "sites", type = "t")
```
Extract NMDSRun1 Scores.
```{r}
NMDSRun1Scores <- scores(NMDSRun1, display = "sites")
```

##Visualizing NMDS Data

Extract Tree Canopy Health Scores.
```{r}
TreeCanopyHealthScoringPrep <- CombinedCleanedMatrix[,c("field.tree.health.condition")]
```

Visualizing data with color scores.
```{r}
ggplot(NMDSRun1Scores, aes(x = NMDS1, y = NMDS2, color = TreeCanopyHealthScoringPrep)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Good" = "forestgreen", "Fair" = "yellow", "Poor" = "orange", "Dead" = "red")) +
  theme_minimal() +
  labs(title = "NMDS with Tree Health Scores", color = "Tree Health") +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-1, 1))
```
As of the day of writing this after running the final NMDS analysis it seems we just don't have a large enough sample size with enough dead trees to confirm if there is any relationship between understory plants and overall tree health. As this project continues to grow with more data collected and more parks assessed future assessments on the data may yield more interesting results.


##Sorting Tree Condition

Separating column for Tree Conditions.
```{r}
good_count <- TreeCanopyHealthData %>%
  summarise(good_count = sum(field.tree.health.condition == "Good")) %>%
  pull(good_count)
fair_count <- TreeCanopyHealthData %>%
  summarise(fair_count = sum(field.tree.health.condition == "Fair")) %>%
  pull(fair_count)
poor_count <- TreeCanopyHealthData %>%
  summarise(poor_count = sum(field.tree.health.condition == "Poor")) %>%
  pull(poor_count)
dead_count <- TreeCanopyHealthData %>%
  summarise(dead_count = sum(field.tree.health.condition == "Dead")) %>%
  pull(dead_count)
```

Create list of Tree Conditions.
```{r}
tcs=c("602", "128", "28", "1")
```

Create from list of tree conditions data frame on tree condition.
```{r}
Tree.Condition.Summary <- data.frame(
  category=c("Good","Fair","Poor", "Dead") , 
  value=as.numeric(tcs)
)
```

Bar chart.
```{r}
ggplot(Tree.Condition.Summary, aes(x=reorder(category, -value) , y=value)) +
  geom_bar(stat = "identity") +
  labs(x = "Tree Condition", y = "Tree Count", title = "Tree Condition of Western Red Cedar Found at Point Defiance 2025")
```

##Sorting DBH Sizes into Buckets

For tree size a histogram with bucket size **5 inch** for now should suffice for our purposes.

Separate DBH column into its own data frame.
```{r}
DBH.Inch <- i.nat.all[c(12)]
```

Create histogram. (Bin = 5 Inches starting from 0)
```{r}
ggplot(DBH.Inch, aes(x = field.tree.diameter)) + 
  geom_histogram(binwidth = 5, boundary = 0, fill = "springgreen4", color = "black") + 
  stat_bin(binwidth = 5, boundary = 0, geom = "text", aes(label = after_stat(count), vjust = -0.5)) + 
  labs(x = "DBH", y = "Count", title = "Western Red Cedar Tree Sizes Found at Point Defiance 2025")
```

The 27 entries removed were entries without DBH's entered.

##ArcGIS Species Hotspots Mapping GIS Prep

In order to create the species hot spots maps I will need to attach GPS coordinates from the i.nat.all dataframe with the SpeciesMatrix dataframe and then filter for presence from the 1s and 0s from each column and separate each of those into CSV files for ArcGIS.

Creating dataframe with just longitude, latitude and id.
```{r}
GPSCoords <- i.nat.all[c(1,3,4)]
```

Combining based off of ID columns in GPSCoords with SpeciesMatrix.
```{r}
GPSCoordPrep <- left_join(SpeciesMatrix,GPSCoords, by = "id")
```

Filtering rows based on presence of species X and turning created dataframes into CSVs.

Salal.
```{r}
GisPrepSalal <- GPSCoordPrep %>%
  filter(`salal` == 1) %>%
  select(`salal`, latitude, longitude)
write_csv(GisPrepSalal, "Salal.csv")
```

Wax Myrtle.
```{r}

```

Fern.
```{r}

```

Ivy.
```{r}

```

Stinking Iris.
```{r}

```

Huckleberry.
```{r}

```

Cedar.
```{r}

```

Holly.
```{r}

```

Sword Fern.
```{r}

```

Horsetail.
```{r}

```

Bracken Fern.
```{r}

```

Cotoneaster.
```{r}

```

Nettle.
```{r}

```

Salmon Berry.
```{r}

```

Cherry Laurel.
```{r}

```

Oregon Grape.
```{r}

```

Trailing Blackberry.
```{r}

```

Ocean Spray.
```{r}

```

Evergreen Huckleberry.
```{r}
GisPrepEvergreenHuckleberry <- GPSCoordPrep %>%
  filter(`evergreen huckleberry` == 1) %>%
  select(`evergreen huckleberry`, latitude, longitude)
write_csv(GisPrepEvergreenHuckleberry, "Evergreen Huckleberry.csv")
```

Western Starflower.
```{r}

```

Himalayan Blackberry.
```{r}

```

Horse Chestnut.
```{r}

```

Pacific Sanicle.
```{r}

```

American Trailplant.
```{r}

```

Mountain Sweet Cicely.
```{r}

```








